### 2024-03-28 manuscript/manuscript_blinded.qmd
anonymize GitHub links

#### **Childhood adversity is not associated with lowered inhibition, but slower perceptual processing: A Drift Diffusion Model analysis**

#### Stefan Vermeent^a,b^, Ethan S. Young^a^, Jean-Louis van Gelder^b^, & Willem E. Frankenhuis^a,b,c^

#### ^a^ Department of Psychology, Utrecht University, Utrecht, The Netherlands

#### ^b^ Max Planck Institute for the Study of Crime, Security, and Law, Freiburg, Germany

#### ^c^ Evolutionary and Population Biology, Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Amsterdam, the Netherlands

# Corresponding author

Correspondence concerning this article should be addressed to Stefan Vermeent, Department of Psychology, Utrecht University, Heidelberglaan 1, 3584 CS Utrecht, The Netherlands. Email: p.c.s.vermeent@gmail.com / p.c.s.vermeent@uu.nl.

# Data Statement

All scripts, materials, and data supporting the results presented here are available on the article's GitHub repository [ANONYMIZED FOR REVIEW].

# Funding Statement
WEF’s contributions have been supported by the Dutch Research Council (V1.Vidi.195.130) and the James S. McDonnell Foundation (https://doi.org/10.37717/220020502).

# Author contributions

**Stefan Vermeent:** Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Project administration, Data curation, Visualization, Writing - original draft, Writing - review & editing.
**Ethan S. Young:** Conceptualization, Methodology, Supervision, Writing - review & editing.
**Jean-Louis van Gelder:** Conceptualization, Funding acquisition, Methodology, Supervision, Writing - review & editing.
**Willem E. Frankenhuis:** Conceptualization, Funding acquisition, Methodology, Supervision, Writing - review & editing.

@@

Previous research has shown that people who grew up in adverse conditions tend to be slower on the Flanker Task, suggesting difficulty inhibiting distractions.
However, this pattern might result from slower general cognitive processes (e.g., reduced general processing speed), rather than the specific ability of inhibition.
We used Drift Diffusion Modeling in three studies---one pilot and two confirmatory studies---to better understand associations of adversity exposures with Flanker performance. 
Neither exposure to violence nor unpredictability were associated with inhibition.
In contrast, although mixed, exposure to violence and unpredictability were associated with lower strength of perceptual input; that is, how well someone can process the arrows in general.
Finally, lower strength of perceptual input was linked to a tendency to process information more holistically, focusing less on individual details.
Thus, researchers should be cautious when drawing inferences about inhibition abilities in people exposed to adversity based on response times alone.

@@

We used an incremental preregistration approach across studies (for all preregistrations, code and materials, see [ANONYMIZED FOR REVIEW].
For each study, we preregistered primary (i.e., hypothesis-driven) and exploratory analyses.
The main text addresses the primary analyses involving violence exposure and the exploratory analyses involving environmental unpredictability.
We describe the other exploratory analyses in the supplemental materials (section 2).
For an overview of all deviations from the preregistrations, see section 4 of the supplemental materials.

@@

Participants were 565 US-based individuals aged between 18 and 30 recruited on Prolific Academic ([https://www.prolific.co](https://www.prolific.co)) (See Table 1 for demographic data).
The sample was balanced on sex.
We used the MacArthur's ladder, included in Prolific's prescreening battery, for assessing perceived SES to ensure about half of the sample came from lower-SES backgrounds (which we defined as a score of 4 or below).
Participants were eligible if they spoke fluent English and did not report color-blindness.

@@

We conducted a power simulation using the *faux* package in R [@debruine_2021] to determine the minimally required number of participants for standardized regression coefficients of 0.10 and 0.15 (for details and simulation code, see [ANONYMIZED FOR REVIEW]).
Power was \> .80 for adversity x task condition interactions with *N* = 450 or more.
For a linear main effect, detecting an effect of $\beta$ = 0.15 with .90 power would require *N* = `r pwr.r.test(r = 0.15, sig.level = 0.05, power = .90, alternative = "two.sided")[['n']] %>% round(0)`.
We sampled 550 participants, anticipating a final sample of \~500 after exclusions.

@@

**Multiverse analysis.** We included the same decisions in the multiverse as in the previous studies.
However, there was one deviation from the preregistration: the multiverse analysis contained the same arbitrary decisions as the Pilot and Study 1, instead of a subset, as we preregistered (for details, [ANONYMIZED FOR REVIEW]).
See the supplemental materials (section 5) for figures summarizing *p*-distributions and the explained variance in the regression coefficients of each data cleaning decision.

@@

**Boundary separation:** Parameter of the Drift Diffusion Model and Shrinking Spotlight Model. Measures the amount of information a person requires before making a decision, determining the speed-accuracy trade-off (e.g., sacrificing speed in favor of higher accuracy).

>>>>

@@

It is well-established that individuals who grew up in adverse conditions tend to be slower on the Flanker Task. 
This finding is typically interpreted to reflect difficulty inhibiting distractions. 
However, it might result from slower general cognitive processes (e.g., reduced general processing speed), rather than the specific ability of inhibition. 
We used Drift Diffusion Modeling in three online studies (total N = 1560) with young adults to understand associations of adversity with Flanker performance. 
We find no associations between exposure to violence and unpredictability with inhibition. 
Yet, although mixed, violence and unpredictability exposure were associated with lower strength of perceptual input--—how well someone can process target and distractor information alike. 
Finally, people with lower strength of perceptual input processed information more holistically, focusing less on details. 
Thus, lowered Flanker performance does not necessarily imply lowered inhibition ability. 
Cognitive modeling might reveal a different picture of abilities in adverse conditions.

@@

We used an incremental preregistration approach across studies (for all preregistrations, code and materials, see <https://anonymous.4open.science/r/CogDev_AnonLink>).
For each study, we preregistered primary (i.e., hypothesis-driven) and exploratory analyses.
The main text addresses the primary analyses involving violence exposure and the exploratory analyses involving environmental unpredictability.
We describe the other exploratory analyses in the supplemental materials (section 2).
For an overview of all deviations from the preregistrations, see section 4 of the supplemental materials.

@@

Participants were 565 people from the United States aged between 18 and 30 recruited on Prolific Academic ([https://www.prolific.co](https://www.prolific.co)) (See Table 1 for demographic data).
The sample was balanced on sex.
We used the MacArthur's ladder, included in Prolific's prescreening battery, for assessing perceived SES to ensure about half of the sample came from lower-SES backgrounds (which we defined as a score of 4 or below).
Participants were eligible if they spoke fluent English and did not report color-blindness.

@@

We conducted a power simulation using the *faux* package in R [@debruine_2021] to determine the minimally required number of participants for standardized regression coefficients of 0.10 and 0.15 (for details and simulation code, see <https://anonymous.4open.science/r/CogDev_AnonLink/preregistrations/1_pilot/scripts>).
Power was \> .80 for adversity x task condition interactions with *N* = 450 or more.
For a linear main effect, detecting an effect of $\beta$ = 0.15 with .90 power would require *N* = `r pwr.r.test(r = 0.15, sig.level = 0.05, power = .90, alternative = "two.sided")[['n']] %>% round(0)`.
We sampled 550 participants, anticipating a final sample of \~500 after exclusions.

@@

**Multiverse analysis.** We included the same decisions in the multiverse as in the previous studies.
However, there was one deviation from the preregistration: the multiverse analysis contained the same arbitrary decisions as the Pilot and Study 1, instead of a subset, as we preregistered (for details, <https://anonymous.4open.science/r/CogDev_AnonLink/preregistrations/3_study2>).
See the supplemental materials (section 5) for figures summarizing *p*-distributions and the explained variance in the regression coefficients of each data cleaning decision.

@@

**Boundary separation:** Parameter of the Drift Diffusion Model and Shrinking Spotlight Model. Measures the amount of information a person requires before making a decision (i.e., response caution).
